The Right Way to Search Evolving Graphs

### 2016-03-04 paper.tex
Final round of edits

\begin{abstract}
Evolving graphs arise in problems where interrelations between data change over
time. We present a breadth first search (BFS) algorithm for evolving graphs
that computes the most direct influences between nodes at two different times.
Using simple examples, we show that na\"ive unfoldings of adjacency matrices
miscount the number of temporal paths.
By mapping an evolving graph to an adjacency matrix of an equivalent static graph,
we prove that our generalization of the BFS algorithm is correct.
Finally, we demonstrate how the BFS over evolving graphs can be applied to mine
citation networks.

@@

Treatments of evolving graphs vary in their generality and focus.
Kivel\"a \textit{et al.}~\cite{kabg14} treat time dependence as a special
case of families of graphs with multiple interrelationships.
Others like Flajolet \textit{et al.}~\cite{fkp89} use time to index the family
of related graphs, but are not concerned with explicit time-dependent processes.
Yet others focus on incremental updates to large graphs~\cite{bkmu12}.
Here, we describe evolving graphs as a time-ordered sequence of graphs, similar
to the study of metrized graphs by Tang and coworkers~\cite{ntmm13,tmml09,tsmm09,tmml10} and of
the dynamics of communities by Grindrod, Higham and coworkers~\cite{gphe11,grihig13}.

The game described in the beginning can be encoded in an evolving graph.
The spread of information to the winner can be described in terms of traversing
this graph using discrete paths that step in both space and time.
Traversals of an ordinary (static) graph may be computed using well known methods
such as the breadth-first search (BFS). An informal description of BFS generalized
to evolving graphs can be found in \cite{tmml09}.
However, it turns out that na\"ive extensions can lead to incorrect descriptions of
the resulting graph traversals. A proper treatment requires the notion of
\emph{node activeness} to describe the set of paths that can only traverse time
or edges, which we call \textit{temporal paths}.
As a result, our treatment can be applied to any evolving graph, even those that
are highly dynamic with arbitrary changes to the nodes and edges.

It is well known that sparse matrix-vector product is equivalent to a
one-step BFS on the corresponding graph~\cite[Sec. 1.1]{kegi11}.
However, little is known about the generalization of this duality for evolving graphs.
We prove the generalization of this duality by finding
the correct block sparse matrix representation of an evolving graph.

In Section~\ref{sec:breadth-first-search}, we explain how the BFS algorithm can be applied correctly on
evolving graphs. Section~\ref{sec:temporal-paths} provides an example showing that considering
only products of the time-dependent adjacency matrices can lead to incorrect enumeration
of temporal paths. We present and demonstrate the BFS algorithm over evolving graphs
in Section~\ref{sec:evolving-graph-bfs}, showing that it is formally equivalent to
BFS over a particular static graph generated by connecting
together active notes of the evolving graph. This static graph generates an algebraic
representation of the BFS as power iteration of its adjacency matrix to a starting
search node, as shown in Section~\ref{sec:bfsla}. The algebraic formulation also
demonstrates interesting connections between properties of the BFS algorithm and
the adjacency matrix. Finally in Section~\ref{sec:applications},
we explain how BFS on evolving graphs may be applied to study dynamical processes
over citation networks.

@@

The simple example of Figure~\ref{fig:eg_shortest_path} provides several
counterexamples which demonstrate why sums over products of adjacency matrices of
the form \eqref{eq:sum} do not count temporal paths correctly. The interpretation
from considering powers of $A$ for static graphs therefore does not extend to
evolving graphs, and sums like \eqref{eq:sum} cannot form the basis of correct
analyses of evolving graphs as they neglect the combinatorics associated with
the causal edge set $E'$.

@@

  \foreach \from/\to in {n10/n7, n1/n4, n11/n12}
   \draw[every edge,->] (\from) -- (\to);
     \draw[dotted,->](n7) to[out=80, in=40] (n11);
  \draw[dotted,->](n10) to[out=10, in=180] (n1);
   \draw[dotted,->](n4) to[out=-30,in=-150] (n12);
    \end{tikzpicture}
\end{center}
\caption{Static graphs corresponding to the evolving graph example of
Figure~\ref{fig:eg_shortest_path}. The green nodes are active nodes while the
red nodes are inactive nodes.
The black lines are edges in $\cup_t E^{[t]}$ and are encoded
algebraically in the diagonal blocks $A^{[t]}$ of the adjacency matrix $\bm A_3$
or $\bm M_3$.
The dotted lines are edges in $E'$ and are encoded algebraically in
the off-diagonal blocks $M^{[t_i, t_j]}$.
The static graph $G$ constructed in the proof of Theorem~\ref{thm:correct} is
formed by retaining all the edges shown and only the active nodes, and has the
adjacency matrix $\bm A_3$. The graph containing all the edges and temporal nodes
shown has adjacency matrix $\bm M_3$.}
\label{fig:static}
\end{figure}

For the example of Figure~\ref{fig:eg_shortest_path}, we have
\begin{align*}
V & = \{(1,\!t_1), (2,\!t_1), (1,\!t_2), (3,\!t_2), (2,\!t_3), (3,\!t_3)\},\\
\cup_t\!E^{[t]} &= \{((1,\!t_1), (2,\!t_1)), ((1,\!t_2), (3,\!t_2)), ((2,\!t_3), (3,\!t_3))\},\\
E'  &= \{((1,\!t_1), (1,\!t_2)), ((2,\!t_2), (2,\!t_3)), ((3,\!t_2), (3,\!t_3))\}.
\end{align*}
In the order specified for $V$, the adjacency matrix of $G$ is then
\[
\bm A_3 = \begin{bmatrix}
0 & 1 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 0
\end{bmatrix}
\]

@@

We see that $(\bm A_n^T)^2$ correctly encodes the collection of products in
\eqref{eq:5}-\eqref{eq:6}. Furthermore, $(\bm A_3^T)^3 \bm b$ correctly counts the
two allowed temporal paths from $(1, t_1)$ to $(3, t_3)$, and that the
off-diagonal structure encoded in $E'$ and the $M^{[t,t']}$ blocks are critical
to obtaining the correct count.

@@

We generate a sequence of random (directed) \texttt{IntEvolvingGraph}s
with $1.0\times 10^5$ active nodes and $10$ time stamps.
The first \texttt{IntEvolvingGraph} in the sequence has about
$1.0 \times 10^8$ static edges.
We consecutively add new
random static edges to this \texttt{IntEvolvingGraph}. For example, 
the second random \texttt{IntEvolvingGraph} in the sequence 
has about $1.5 \times 10^8$ static edges and the third
has $1.8 \times 10^8$ static edges.
% At each time stamp $i$, a static edge is included in $G^{[i]} \in G_4$ with
% probability $p$. 
Note that in this experiment, we do not have direct control
over the full edge set $E$, only the static edge set $\tilde E$. However, the
causal edge set $E'$ can grow at most by $O(\tilde E)$, so it suffices to demonstrate
linear scaling in $|\tilde E|$. 
Figure \ref{fig:time} shows the plots of number of edges against the computation time
for running Algorithm \ref{alg:bfs} in Julia. All experiments are conducted
on a Linux system with Intel(R) Xeon(R) CPU E7-8850 @ $2.00$GHz.
The results show Algorithm \ref{alg:bfs} can be computed in linear time, which agrees
with the result of Theorem \ref{thm:complexity}.

>>>>

\begin{abstract}
Evolving graphs arise in problems where interrelations between data change over
time. We present a breadth first search (BFS) algorithm for evolving graphs
that computes the most direct influences between nodes at two different times.
Using simple examples, we show that na\"ive unfoldings of adjacency matrices
miscount the number of temporal paths.
By mapping an evolving graph to an adjacency matrix of an equivalent static graph,
we prove that our generalization of the BFS algorithm correctly accounts for
paths that traverse both space and time.
Finally, we demonstrate how the BFS over evolving graphs can be applied to mine
citation networks.

@@

Treatments of evolving graphs vary in their generality and focus.
Kivel\"a \textit{et al.}~\cite{kabg14} treat time dependence as a special
case of families of graphs with multiple interrelationships.
Others like Flajolet \textit{et al.}~\cite{fkp89} use time to index the family
of related graphs, but are not concerned with explicit time-dependent processes.
Yet others focus on incremental updates to large graphs~\cite{bkmu12}.
Here, we describe evolving graphs as a time-ordered sequence of graphs, similar
to the study of metrized graphs by Tang and coworkers~\cite{ntmm13,tmml09,tsmm09,tmml10} and of
community dynamics by Grindrod, Higham and coworkers~\cite{gphe11,grihig13}.

The game described above can be encoded in an evolving graph.
The spread of information to the winner can be described in terms of traversing
this graph using discrete paths that step in both space and time.
Traversals of an ordinary (static) graph may be computed using well known methods
such as the breadth-first search (BFS). An informal description of BFS generalized
to evolving graphs can be found in \cite{tmml09}.
However, it turns out that na\"ive extensions can lead to incorrect descriptions of
the resulting graph traversals by accounting for traversals of edges in space,
but not necessarily in time. A proper treatment requires the notions of
\emph{node activeness} to describe the set of paths that can only traverse time
or edges, which we call \textit{temporal paths}, as well as \emph{causal edges}
which connect active nodes with the same identity across different times.
As a result, our treatment can be applied to any evolving graph, even those that
are highly dynamic with arbitrary changes to the nodes and edges.

It is well known that sparse matrix-vector product is equivalent to a
one-step BFS on the corresponding (static) graph~\cite[Sec. 1.1]{kegi11}.
In this paper, we demonstrate a correct corresponding result for an evolving
graph by constructing a block triangular matrix representation of the graph that
takes into account both static and causal connections between active nodes.
In Section~\ref{sec:breadth-first-search}, we explain how the BFS algorithm can
be applied to evolving graph to enumerate paths that traverse edges across time
and space. Section~\ref{sec:temporal-paths} provides an example showing that considering
only products of the time-dependent adjacency matrices fails to enumerate certain
temporal paths. We present and demonstrate the BFS algorithm over evolving graphs
in Section~\ref{sec:evolving-graph-bfs}, showing its formal equivalence to
BFS over a particular static graph generated by adding causal edges that connect
active nodes. This static graph generates an algebraic
representation of the BFS as power iteration of its adjacency matrix to a starting
search node, as shown in Section~\ref{sec:bfsla}. The algebraic formulation also
demonstrates interesting connections between properties of the BFS algorithm and
the adjacency matrix. We describe in Section~\ref{sec:implementation-julia} an
implementation of the algorithm in Julia. Finally in Section~\ref{sec:applications},
we explain how BFS on evolving graphs may be applied to study dynamical processes
over citation networks.

@@

The simple example of Figure~\ref{fig:eg_shortest_path} demonstrates why sums
over products of adjacency matrices of the form \eqref{eq:sum} do not count
temporal paths correctly: they neglect the combinatorics associated with
the causal edge set $E'$. In the next section, we show how to account for these
causal edges by introducing a new matrix--vector product $\ocirc$.

@@

  \foreach \from/\to in {n10/n7, n1/n4, n11/n12}
   \draw[every edge,->] (\from) -- (\to);
     \draw[dotted,->](n7) to[out=80, in=40] (n11);
  \draw[dotted,->](n10) to[out=10, in=180] (n1);
   \draw[dotted,->](n4) to[out=-30,in=-150] (n12);
    \end{tikzpicture}
\end{center}
\caption{Static graphs corresponding to the evolving graph example of
Figure~\ref{fig:eg_shortest_path}. The green nodes are active nodes while the
red nodes are inactive nodes.
The black lines are edges in the static edge set $\tilde E$ and are encoded
algebraically in the diagonal blocks $A^{[t]}$ of the adjacency matrix $\bm A_3$
or $\bm M_3$.
The dotted lines are edges in the causal edge set $E'$ and are encoded algebraically in
the off-diagonal blocks $M^{[t_i, t_j]}$.
The static graph $G$ constructed in the proof of Theorem~\ref{thm:correct} is
formed by retaining all the edges shown and only the active nodes, and has the
adjacency matrix $\bm A_3$. The graph containing all the edges and temporal nodes
shown has adjacency matrix $\bm M_3$.}
\label{fig:static}
\end{figure}

For the example of Figure~\ref{fig:eg_shortest_path}, we have
\begin{align*}
V & = \{(1,\!t_1), (2,\!t_1), (1,\!t_2), (3,\!t_2), (2,\!t_3), (3,\!t_3)\},\\
\tilde E &= \{((1,\!t_1), (2,\!t_1)), ((1,\!t_2), (3,\!t_2)), ((2,\!t_3), (3,\!t_3))\},\\
E'  &= \{((1,\!t_1), (1,\!t_2)), ((2,\!t_2), (2,\!t_3)), ((3,\!t_2), (3,\!t_3))\}.
\end{align*}
In the order specified for $V$, the adjacency matrix of $G$ is then
\[
\bm A_3 = \begin{bmatrix}
0 & 1 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 1 \\
0 & 0 & 0 & 0 & 0 & 0
\end{bmatrix}
\]

@@

We see that $(\bm A_n^T)^2$ encodes all the products in
\eqref{eq:5}-\eqref{eq:6}, including both the ordinary matrix--vector product
and the $\ocirc$ product. Furthermore, $(\bm A_3^T)^3 \bm b$ correctly counts the
two allowed temporal paths from $(1, t_1)$ to $(3, t_3)$, and that the
off-diagonal structure encoded in $E'$ and the $M^{[t,t']}$ blocks are critical
to obtaining the correct count.

@@

We generate a sequence of random (directed) \texttt{IntEvolvingGraph}s
with $10^5$ active nodes and $10$ time stamps.
The first \texttt{IntEvolvingGraph} in the sequence has about
$10^8$ static edges.
We consecutively add new
random static edges to this \texttt{IntEvolvingGraph}. For example,
the second random \texttt{IntEvolvingGraph} in the sequence
has about $1.5 \times 10^8$ static edges and the third
has $1.8 \times 10^8$ static edges.
% At each time stamp $i$, a static edge is included in $G^{[i]} \in G_4$ with
% probability $p$.
Note that in this experiment, we do not have direct control
over the full edge set $E$, only the static edge set $\tilde E$. However, the
causal edge set $E'$ can grow at most by $O(\tilde E)$, so it suffices to demonstrate
linear scaling in $|\tilde E|$.
Figure \ref{fig:time} shows the plots of number of edges against the computation time
for running Algorithm \ref{alg:bfs} in Julia. All experiments are conducted
on a single core of a Linux system with an Intel(R) Xeon(R) E7-8850 CPU running
at 2.00 GHz clock speed.
The results show Algorithm \ref{alg:bfs} can be computed in linear time, which agrees
with the result of Theorem \ref{thm:complexity}.
